---
title: "Webscraping NBA data: 2015-16 Golden State Warriors vs 1995-96 Chicago Bulls"
author: "Danny Wong"
date: "`r format(Sys.time(), '%d %B, %Y')`"
layout: post
blog: true
tag:
- R
- coding
---

The 2015-16 Golden State Warriors are on the verge of either equalling or beating the 1995-96 Chicago Bulls' record of 72 wins in the NBA regular season. Let's use `r` to scrape some data from NBA.com to graph the Warriors' progress over the season in comparison to the Bull's record-setting season.

We will use the `rvest` [package](http://blog.rstudio.org/2014/11/24/rvest-easy-web-scraping-with-r/) to do so.

```{r, warning=FALSE}
library(rvest)
library(tidyr)

nba <- html("http://www.nba.com/news/2015-16-golden-state-warriors-chase-1995-96-chicago-bulls-all-time-wins-record/")
data <- nba %>% html_nodes("table") %>% html_table(header = TRUE, fill = TRUE)

summary(data)

data <- data[[2]]

head(data)

#split the dataframes into seperate Warriors and Bulls dataframes 
colnames(data)[3] <- "Game" #To remove the space between Game Number

warriors <- cbind(data[3], data[2])
bulls <- cbind(data[3], data[5])

#Separate the wins and losses into 2 columns
warriors <- warriors %>% separate(col = Record, into = c("W", "L"), sep = "-")
bulls <- bulls %>% separate(col = Record, into = c("W", "L"), sep = "-")

#Plot with base graphics
plot(W ~ Game, type="l", col="red", data = bulls, lty = 2) #Dotted red line for Bulls
points(W ~ Game, type="l", col="blue", data = warriors) #Solid blue line for Warriors 
```

Do we think that the Warriors can get to 73 wins this season? Let's try and predict this with a linear regression model, now for a bit of fun! Of course this is fraught with methodological problems, so don't gamble any money based on the result.
`r warriors[80:82, 2] <- NA`

```{r}
#Fit a linear model
#Add a regression line
model <- lm(W ~ Game, data = warriors)
summary(model)

plot(W ~ Game, type="l", col="red", data = bulls, lty = 2)
points(W ~ Game, type="l", col="green", data = warriors)

#Add a dotted black line for the regression line
abline(model, lty = 2)

#Predict number of wins
#Add a regression line
predict(model, newdata = data.frame(Game = 82), interval = "prediction")
```

Ok the model says that the Warriors will have **`r predict(model, newdata = data.frame(Game = 82), interval = "prediction")[1] %>% signif(2)`** wins this season. BUT, they now have **70** wins and only **3** games left, making it an impossibility to reach 74 wins. So predicting based purely on this model can't be right.