---
title: "Geographical variation in Critical Care bed capacity (Part 2)"
author: "Danny Wong"
date: "10 December, 2016"
layout: post
blog: true
tag:
- R
- Coding
---

In my [last post](http://dannyjnwong.github.io/Geographical-variation-in-Critical-Care-bed-capacity/), I started cleaning some openly accessible data to look at whether we can see if there was geographical variation in Critical Care beds. 

We ended up with a couple of choropleths which didn't look that great. There were large areas where there were no Critical Care beds, and part of the problem is the CCG geography we chose to map the beds against. This time I'm going to try and map it against English Ceremonial Counties. The [shapefile for this is available from data.gov.uk](https://data.gov.uk/dataset/ceremonial-county-boundaries-of-england).

```{r, message=FALSE, warning=FALSE}
#Load the required packages
library(dplyr)
library(readxl)
library(rgdal)
library(sp)
library(ggplot2)
```

Assuming we have already loaded the data we needed from the previous post. We just need to load the shapefile for English Ceremonial Counties.

```{r, message=FALSE, warning=FALSE, include=FALSE}
#Load the data
CC_beds <- read_excel("../data/MSitRep-October-2016-1VHw9.xls", skip = 14)[3:155,2:4] %>%
  rename(beds = `Number of Adult critical care beds`)

#Load the trust data with Latitude and Longitude coordinates
trust <- read.csv("../data/trust.csv")

#Join the two to get the latitude and longitude codes
CC_beds <- left_join(CC_beds, trust, by = c("Code" = "OrganisationCode" )) %>%
  select(Code, HospitalName = Name, beds, Latitude, Longitude)

#There are some Trusts with NA values for Latitude and Longitude
#This is likely because the Trusts operate over a number of sites
#To get around this we have to use different Hospital data and get the geometric mean coordinates
hospitals <- read.csv("../data/Hospital 20160921.csv")
x <- CC_beds %>% filter(is.na(Latitude)) %>%
  left_join(hospitals, by = c("Code" = "ParentODSCode")) %>%
  select(Code, HospitalName, beds, Latitude = Latitude.y, Longitude = Longitude.y) %>% 
  group_by(Code) %>% 
  summarise(Latitude = mean(Latitude), Longitude = mean(Longitude))
CC_beds <- CC_beds %>% filter(is.na(Latitude)) %>%
  select(-Latitude, -Longitude) %>%
  left_join(x, by = "Code") %>%
  rbind(CC_beds %>% filter(!is.na(Latitude)))

#Join the coordinates to the CC_beds

CCG_pop <- read_excel("../data/SAPE18DT5-mid-2015-ccg-syoa-estimates.xls", skip = 4, sheet = 2)[c(7:18, 21:31, 34:53, 56:78, 83:96, 99:113, 116:133, 136:149, 154:185, 190:198, 201:214, 217:236, 239:245),c(1, 4:5)]
colnames(CCG_pop) <- c("code", "CCG", "pop")

IMD_CCG <- read_excel("data/File_13_ID_2015_Clinical_Commissioning_Group_Summaries.xlsx", sheet = 2)
colnames(IMD_CCG) <- c("Code", "Name", "IMD_avg_rank", "IMD_rank_avgrank", "IMD_avg_score", "IMD_rank_avg_score", "IMD_prop_LSOA", "IMD_rank_prop_LSOA", "IMD_extent", "IMD_rank_extent", "IMD_local_conc", "IMD_rank_local_conc")

```

```{r, message=FALSE, warning=FALSE}
counties_map <- readOGR(dsn = "../data/English Ceremonial Counties", layer ="English Ceremonial Counties")
```

Again we confirm that the polygons have loaded correctly by plotting quickly.

```{r}
plot(counties_map)
points(x = CC_beds$Longitude, y = CC_beds$Latitude, col = "red", pch = 19)
```

Like the last time, the next step would be to do a spatial join. To find out how many Critical Care beds there are in each county.

```{r, message=FALSE, warning=FALSE}
CC_beds_sp <- CC_beds
coordinates(CC_beds_sp) <- ~ Longitude + Latitude
proj4string(CC_beds_sp) <- CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
CC_beds <- cbind(CC_beds, over(CC_beds_sp, map)) 
Counties_CC_beds_join <- CC_beds %>% group_by(NAME) %>% 
  summarise(sum(beds)) %>%
  rename(crit_care_beds = `sum(beds)`)

nrow(Counties_CC_beds_join)
head(Counties_CC_beds_join)
tail(Counties_CC_beds_join)
```

Let's now plot a choropleth!

```{r, message=FALSE, warning=FALSE}
#We will first need to fortify the map SpatialPolygonsDataframe so that ggplot2 can plot it
#We need to use gBuffer, because there's a problem with the shapefile with self-intersecting
#Solution to this problem is from: http://gis.stackexchange.com/questions/163445/r-solution-for-topologyexception-input-geom-1-is-invalid-self-intersection-er
counties_map.f <- fortify(rgeos::gBuffer(counties_map, byid=TRUE, width=0), region = "NAME")

#Now let's merge the Critical Care Beds data
merge.counties_map.f <- merge(counties_map.f, Counties_CC_beds_join, by.x = "id", by.y = "NAME", all.x=TRUE) #%>% 
  #mutate(crit_care_beds_per_pop = replace(crit_care_beds_per_pop, which(is.na(crit_care_beds_per_pop)), 0)) %>%
  #mutate(crit_care_beds = replace(crit_care_beds, which(is.na(crit_care_beds)), 0))

#Reorder otherwise the plot will look weird
final.plot <- merge.counties_map.f[order(merge.counties_map.f$order), ] 

#Plot!
ggplot() +
  geom_polygon(data = final.plot, aes(x = long, y = lat, group = group, fill = crit_care_beds),
               color = "black", size = 0.1) +
  coord_map() +
  theme_minimal()
```

This is certainly a more pleasing map than the previous two. However, there is a problem: we don't have a population-weighted representation. Getting population estimates is tedious. Fortunately someone has already done this [online, based on the 2014 census estimates (Wikipedia)](https://en.wikipedia.org/wiki/List_of_ceremonial_counties_of_England).

```{r}
county_pop <- read.csv("../data/county_populations_2014.csv")
Counties_CC_beds_join <- Counties_CC_beds_join %>% full_join(county_pop, by = c("NAME" = "County")) %>%
  mutate(crit_care_beds_per_pop = crit_care_beds/Population * 100000)

merge.counties_map.f <- merge(counties_map.f, Counties_CC_beds_join, by.x = "id", by.y = "NAME", all.x=TRUE) 
final.plot <- merge.counties_map.f[order(merge.counties_map.f$order), ] 

#Plot!
ggplot() +
  geom_polygon(data = final.plot, aes(x = long, y = lat, group = group, fill = crit_care_beds_per_pop),
               color = "black", size = 0.1) +
  coord_map() +
  theme_minimal()
```

```{r}
sessionInfo()
```