---
title: "Geographical variation in Critical Care bed capacity (Part 1)"
author: "Danny Wong"
date: "09 December, 2016"
layout: post
blog: true
tag:
- R
- Coding
---

I've always wondered if there was variation across the UK with regards to Critical Care bed capacity. In other words, are there regions in the UK where there are more Critical Care beds per population than others?

There're publically available datasets which we can use to try and answer this question:

- NHS England hosts data on the [Critical Care capacity by month and by trust](https://www.england.nhs.uk/statistics/statistical-work-areas/critical-care-capacity/critical-care-bed-capacity-and-urgent-operations-cancelled-2016-17-data/).
- The Office of National Statistics hosts estimates on the [population sizes in the different Clinical Commissioning Groups](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/clinicalcommissioninggroupmidyearpopulationestimates).
- NHS Choices hosts a [list of all the Trusts and Hospitals in England](http://www.nhs.uk/aboutNHSChoices/aboutnhschoices/how-we-perform/Pages/datasets.aspx) with their Latitude and Longitude codes.
- Lastly NHS England hosts [a .kml file for all the CCG regions](https://www.england.nhs.uk/resources/ccg-maps/), which we would need to join the data together by region. Unfortunately it's difficult to read `.kml` files into R, and there're workarounds to do this, but I decided to convert `.kml` file into a `.GeoJSON` file using QGIS.

```{r, message=FALSE, warning=FALSE}
#Load the required packages
library(dplyr)
library(readxl)
library(rgdal)
library(sp)
library(ggplot2)

#Load the data
CC_beds <- read_excel("../data/MSitRep-October-2016-1VHw9.xls", skip = 14)[3:155,2:4] %>%
  rename(beds = `Number of Adult critical care beds`)

#Load the trust data with Latitude and Longitude coordinates
trust <- read.csv("../data/trust.csv")

#Join the two to get the latitude and longitude codes
CC_beds <- left_join(CC_beds, trust, by = c("Code" = "OrganisationCode" )) %>%
  select(Code, HospitalName = Name, beds, Latitude, Longitude)

#There are some Trusts with NA values for Latitude and Longitude
#This is likely because the Trusts operate over a number of sites
#To get around this we have to use different Hospital data and get the geometric mean coordinates
hospitals <- read.csv("../data/Hospital 20160921.csv")
x <- CC_beds %>% filter(is.na(Latitude)) %>%
  left_join(hospitals, by = c("Code" = "ParentODSCode")) %>%
  select(Code, HospitalName, beds, Latitude = Latitude.y, Longitude = Longitude.y) %>% 
  group_by(Code) %>% 
  summarise(Latitude = mean(Latitude), Longitude = mean(Longitude))
CC_beds <- CC_beds %>% filter(is.na(Latitude)) %>%
  select(-Latitude, -Longitude) %>%
  left_join(x, by = "Code") %>%
  rbind(CC_beds %>% filter(!is.na(Latitude)))

#Join the coordinates to the CC_beds

CCG_pop <- read_excel("../data/SAPE18DT5-mid-2015-ccg-syoa-estimates.xls", skip = 4, sheet = 2)[c(7:18, 21:31, 34:53, 56:78, 83:96, 99:113, 116:133, 136:149, 154:185, 190:198, 201:214, 217:236, 239:245),c(1, 4:5)]
colnames(CCG_pop) <- c("code", "CCG", "pop")

#Load the map
#Make sure you have the right OGR drivers installed by typing "GeoJSON" %in% ogrDrivers()$name
map <- readOGR("../data/CCG_BSC_Apr2015.geojson", "OGRGeoJSON")
map$Name <- gsub("&", "and", map$Name)
map$Name <- gsub("Airedale, Wharfdale and Craven", "Airedale, Wharfedale and Craven", map$Name)
map$Name <- gsub("South East Staffs and Seisdon Peninsular", "South East Staffordshire and Seisdon Peninsula", map$Name)
map$Name <- gsub("North, East, West Devon", "Northern, Eastern and Western Devon", map$Name)
map$Name <- paste("NHS", map$Name)
map$Name <- toupper(map$Name)
map[which(map$Name == "NHS WEST LONDON (KANDC AND QPP)"),1] <- "NHS WEST LONDON"
CCG_pop$CCG <- gsub("&", "and", CCG_pop$CCG)
CCG_pop$CCG <- toupper(CCG_pop$CCG)
#x <- sp::merge(map, CCG_pop, by.x = "Name", by.y = "CCG")
```

So now we have all the data loaded up. We can have a look at what we have visually:

```{r}
plot(map)
points(x = CC_beds$Longitude, y = CC_beds$Latitude, col = "red", pch = 19)
```

Not very pretty at the moment, but essentially the polygons represent the CCG regions in England, and each red dot represents a Trust with Critical Care beds.

The next step would be to do a spatial join. To find out how many Critical Care beds there are in each CCG region.

```{r, message=FALSE, warning=FALSE}
CC_beds_sp <- CC_beds
coordinates(CC_beds_sp) <- ~ Longitude + Latitude
proj4string(CC_beds_sp) <- CRS("+proj=longlat +ellps=WGS84 +towgs84=0,0,0,0,0,0,0 +no_defs")
CC_beds <- cbind(CC_beds, over(CC_beds_sp, map)) 
CCG_CC_beds_join <- CC_beds %>% group_by(Name) %>% 
  summarise(sum(beds)) %>% 
  left_join(CCG_pop, by = c("Name" = "CCG")) %>%
  rename(crit_care_beds = `sum(beds)`) %>%
  mutate(crit_care_beds_per_pop = crit_care_beds/pop * 100000) # calculate number of beds per 100k pop

nrow(CCG_CC_beds_join)
head(CCG_CC_beds_join)
tail(CCG_CC_beds_join)
```

Some CCGs have zero Critical Care Beds.

Let's now plot a choropleth!

```{r, message=FALSE, warning=FALSE}
#We will first need to fortify the map SpatialPolygonsDataframe so that ggplot2 can plot it
map.f <- fortify(map, region = "Name")

#Now let's merge the Critical Care Beds data
merge.map.f <- merge(map.f, CCG_CC_beds_join, by.x = "id", by.y = "Name", all.x=TRUE) #%>% 
  #mutate(crit_care_beds_per_pop = replace(crit_care_beds_per_pop, which(is.na(crit_care_beds_per_pop)), 0)) %>%
  #mutate(crit_care_beds = replace(crit_care_beds, which(is.na(crit_care_beds)), 0))

#Reorder otherwise the plot will look weird
final.plot <- merge.map.f[order(merge.map.f$order), ] 

#Plot!
ggplot() +
  geom_polygon(data = final.plot, aes(x = long, y = lat, group = group, fill = crit_care_beds),
               color = "black", size = 0.1) +
  coord_map() +
  theme_minimal()

ggplot() +
  geom_polygon(data = final.plot, aes(x = long, y = lat, group = group, fill = crit_care_beds_per_pop),
               color = "black", size = 0.1) +
  coord_map() +
  theme_minimal()
```

```{r}
sessionInfo()
```