---
title: "Linear Regression in R"
author: "Danny Wong"
date: "18 March 2016"
layout: post
---

I recently attended an excellent [Introduction to Regression Analysis course](https://www.ucl.ac.uk/ich/short-courses-events/short-courses-events-publication/stats-regression) run by the [Centre for Applied Statistics Courses](http://www.ucl.ac.uk/ich/short-courses-events/about-stats-courses) at UCL. They taught the course with example output from SPSS, so here I try to replicate their steps in R.

The data they provided is simulated and in an Microsoft Excel (.xls) spreadsheet format and can be found [here](https://github.com/dannyjnwong/dannyjnwong.github.io/blob/master/data/DataSetFile.xls). So let's start by loading it into R.

```{r Load_Data, warning=FALSE, message=FALSE}
#We will use the readxl package to read the data into R
library(readxl)
data <- read_excel("../data/DataSetFile.xls")

#The last 2 rows need to be removed, because they are actually annotations describing the dataset. The first column can be removed because they are just observation numbers.
data <- data[1:42, 2:9]

#We need to tell R that there are some categorical variables
data$female <- as.factor(as.integer(data$female))
data$drugstat <- as.factor(as.integer(data$drugstat))
data$ethnic <- as.factor(as.integer(data$ethnic))
```

So now we have a dataframe of 42 observations and 8 variables:

  - `frcmax` = VmaxFRC(L/s), a measure of small airway function of the baby at birth
  - `female` = 0 (male), 1 (female)
  - `drugstat` = Whether the mother was administered pethidine during labour: 0 (No), 1 (Yes)
  - `heellen` = Crown-Heel length (cm) of the baby
  - `gestag` = Gestational age (weeks) at birth
  - `ethnic` = Ethnicity: 0 (White Caucasian), 1 (Black African), 2 (Asian)
  - `cd4` = CD4 count (per mm^3) of the baby
  - `bwt` = Birthweight (grams) of the baby

We will use regression to answer the question: **"Is there any association between the use of pethidine during labour and VmaxFRC of the baby at birth?"**

We start by examining the variables, getting a general feel of the data.

```{r Exploratory, warning=FALSE, message=FALSE}
library(Hmisc) #We will use a function from this package which contains many useful functions for data analysis
library(car) #We will use this package to run a scatterplot matrix
describe(data)

scatterplotMatrix(~frcmax + heellen + gestag + cd4 + bwt, diagonal = "density", smoother = FALSE, data = data)

stripchart(frcmax ~ drugstat, vertical = TRUE, offset = .5, pch = 19, ylab = "FRC Max", xlab = "Whether Pethidine was given", data = data)

stripchart(frcmax ~ ethnic, vertical = TRUE, offset = .5, pch = 19, ylab = "FRC Max", xlab = "Ethnicity", data = data)

stripchart(frcmax ~ female, vertical = TRUE, offset = .5, pch = 19, ylab = "FRC Max", xlab = "Sex", data = data)
```

We can now try to fit a linear model, our outcome variable of interest is `frcmax`, and we need to include the predictor `drugstat` in our model. But let's first fit this model `frcmax = a + b(bwt)`

```{r Model-fitting, warning=FALSE, message=FALSE}
model1 <- lm(frcmax ~ bwt, data = data)
summary(model1)

plot(frcmax ~ bwt, pch = 19, data = data)
abline(model1)
```

Now let's add a few more predictors and include the `drugstat` variable.

```{r Model-fitting2, warning=FALSE, message=FALSE}
model2 <- lm(frcmax ~ bwt + gestag + factor(ethnic) + drugstat, data = data)
summary(model2)
```

**So what does this mean?**

In this model, the equation is `frcmax = 226.14 + 0.08(bwt) - 6.49(gestag) - 49.91(ethnic = African) - 27.22(ethnic = Asian) - 17.43(drugstat = 1)`.

The Adjusted R^2 is 0.3241, meaning 32.41% of the variability of `frcmax` can be accounted for by the predictors in the model. `bwt` and `ethnic = African` had p-values <0.05, meaning these were significant independent predictors of `frcmax`. For every 1 unit increase in `bwt`, there is a 6.49 unit increase in `frcmax`, and `ethnic = African` resulted in a mean `frcmax` of -46.91 compared to `ethnic = White Caucasian`.

Since the p-value for `drugstat = 1` is 0.2722, it means that after adjusting for the other variables `bwt`, `gestag`, `ethnic` in this model, the null hypothesis is true and there is no statistically significantassociation between the use of pethidine during labour and VmaxFRC of the baby at birth.