---
title: "Trainees and Rotas: an Assignment Problem"
author: "Danny Wong"
date: "31 January 2020"
layout: post
blog: true
tag:
- coding
- R
---

I saw this tweet last night on Twitter:

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">The worst solution is:<br><br>Line A - Person 1<br> B - 2<br> C - 8<br> D - 6<br> E - 4<br> F - 3<br> G - 7<br> H - 5<br><br>This has a mean choice of 6.5 out of 8. <br><br>What&#39;s the best combination? <a href="https://t.co/HKPrn5MizT">https://t.co/HKPrn5MizT</a> <a href="https://t.co/ZeWnCVmHhC">pic.twitter.com/ZeWnCVmHhC</a></p>&mdash; Graham McCracken (@grahamccracken) <a href="https://twitter.com/grahamccracken/status/1222933576347504642?ref_src=twsrc%5Etfw">January 30, 2020</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

[Graham McCracken](https://twitter.com/grahamccracken) raised a problem that is frequently encountered by hospital consultants and junior doctors who are in charge of the rota: fairly assigning rota slots to trainees according to their preferences. Of course this is essentially an example of the [Assignment Problem](https://en.wikipedia.org/wiki/Assignment_problem).

So I decided to have a go at solving this problem using the `lpSolve` package in `R`.

```{r message=FALSE, warning=FALSE}
# Load library
library(lpSolve)

# Set up the Trainees' preferences
prefT <- matrix(c(1, 1, 1, 1, 1, 1, 1, 1,
                  2, 6, 2, 2, 2, 2, 2, 2,
                  3, 8, 8, 6, 5, 6, 7, 8,
                  4, 5, 7, 3, 7, 8, 6, 3,
                  5, 4, 3, 7, 3, 4, 5, 6,
                  6, 3, 6, 5, 4, 3, 4, 5,
                  7, 2, 4, 8, 6, 5, 8, 7,
                  8, 7, 5, 4, 8, 7, 3, 4),
                nrow = 8, ncol = 8, byrow = TRUE,
                dimnames = list(LETTERS[1:8],
                                1:8))
prefT
```

After loading up the `lpSolve` package, we set up the Trainees' (numbered 1 to 8) preferences (`prefT`) for particular rota lines (A to H) in a matrix. After we have set up the preferences, we can use the `lp.assign()` function to perform the match.

```{r message=FALSE, warning=FALSE}
# Run the assignment
matching <- lp.assign(prefT)
matching$solution
```

The `lp.assign()` returns the solution of the match as a matrix. There should be more than 1 permutation of the solution available but in this instance, the function only returns one permutation. We can then view the result.

```{r}
# Make a dataframe of the results in a human interpretable format
Rota <- rownames(prefT)
ix <- round(matching$solution %*% seq_len(ncol(prefT)))
Trainee <- colnames(prefT)[ifelse(ix == 0, NA, ix)]
Preference <- t(prefT)[as.logical(t(matching$solution))]

results <- data.frame(Rota, Trainee, Preference)

# Print the results dataframe
knitr::kable(results)
```

As you can see, 1 Trainee received their 1st choice, 2 received their 2nd choices, and the remaining 5 received their 3rd choices. This gives a mean preference ranking of `r sum(results$Preference)/nrow(results)` for this solution, which is the best result if you were to try every single permutation of Rota and Trainee there is available. For 8 Rota lines and 8 Trainees, the number of possible permutations would be 8! = 40320.

Also, in this example we have 8 trainees for 8 rota lines, but we could presumably also reformulate it for fewer trainees if there are rota gaps.

```{r}
sessionInfo()
```